Model Finding for Recursive Functions in SMT

### 2016-03-30 ijcar2016/body.tex
tuned; shortened

Nunchaku is a new higher-order model finder designed to be integrated with
several proof assistants. The first version, 0.1, was released in January 2016
with support for (co)algebraic datatypes, (co)recursive functions and
(co)inductive predicates. Support for higher-order functions is in the works.
We have developed a preliminary Isabelle frontend and are
planning further frontends for Coq, %HOL4, HOL Light,
the TLA\textsuperscript{+} Proof System, and possibly other proof assistants.

Nunchaku is the spiritual successor to Nitpick for
Isabelle/HOL~\cite{blanchette-nipkow-2010}, but is developed as a standalone
OCaml program, with its own input language. Whereas Nitpick generates a
succession of problems where cardinalities of finite types grow at each step,
Nunchaku translates its input to one first-order logic program that targets
the finite model finding fragment of CVC4, including
(co)algebraic datatypes \cite{reynolds-blanchette-2015-codata}.
Using CVC4 also allows Nunchaku to provide efficient arithmetic
reasoning and to detect unsatisfiability in addition to satisfiability.
We plan to integrate other tools as backends, to exploit the
strengths of competing approaches.
% and makes
% equality reasoning straightforward.

The input syntax was inspired by that other systems based on higher-order
logic (e.g., Isabelle/HOL) and by functional programming languages (e.g.,
OCaml). The following simple problem gives a taste of the syntax:
%
\begin{quote}
\verb$data nat := 0 | Suc nat.$ \\[2\jot]
\verb$pred even : nat -> prop :=$ \\
\verb$  even 0;$ \\
\verb$  forall n. odd n => even (Suc n)$ \\
\verb$and odd : nat -> prop :=$ \\
\verb$  forall n. even n => odd (Suc n).$ \\[2\jot]
\verb$val m : nat.$ \\
\verb$goal even m && ~ (m = 0).$
\end{quote}
%
The problem defines a datatype (\texttt{nat}) and two mutually recursive
inductive predicates (\texttt{even} and \texttt{odd}), it declares a constant
\texttt{m}, and specifies a goal to satisfy (``\texttt{m} is even and nonzero'').
Nunchaku quickly finds the following partial model:
%
\begin{quote}
\verb$val m := Suc (Suc 0).$ \\
\verb$val odd := fun x. if x = Suc 0 then true else ?__.$ \\
\verb$val even := fun x. if x = Suc (Suc 0) || x = 0 then true else ?__.$\kern-200mm
\end{quote}
%
The partial model gives sufficient information to the user to evaluate the
goal: ``2 is even if 1 is odd, 1 is odd if 0 is even, and 0 is even.''
Our experience with the Nitpick~\cite{blanchette-nipkow-2010} model finder
is that users are mostly interested in the values assigned to uninterpreted
constants (e.g., \texttt{m}). Occasionally, the models of recursive functions
are instructive in the face of underspecification. A typical example is
the \texttt{head} function that returns the first element of a nonempty list:
%
\begin{quote}
\verb$data list A := Nil | Cons A (list A).$ \\[2\jot]
\verb$rec head : pi A. list A -> A :=$ \\
\verb$  forall y ys. head (Cons y ys) = y.$ \\[2\jot]
\verb$goal ~ (head Nil = 0).$
\end{quote}
%
Nunchaku transforms the definition of \verb$head$ into
%
\begin{quote}
\verb$head xs = match xs with Nil -> head xs | Cons y ys -> y end$
\end{quote}
%
where the unspecified \verb$Nil$ case is expressed via nonterminating
recursion (\verb$head$ \verb$xs$ \verb$=$ \verb$head$ \verb$xs$).
The tool exhibits a model in which \verb$head$ \verb$Nil$ is interpreted as
a nonzero value.

@@

Conceptually, the sequence of transformation is a \relax{two-way pipeline}
built by composing pairs $(\textsf{encode},\vthinspace \textsf{decode})$ of
transformations. For each such pair, \textsf{encode} transforms a problem over
a signature $\Sigma$ in a logic~$\mathcal{L}$ to a problem over a signature
$\Sigma'$ in a logic~$\mathcal{L}'$, and \textsf{decode} translates a model
in~$\mathcal{L}'$ over $\Sigma'$ into a model in~$\mathcal{L}$ over $\Sigma$,
in the spirit of institution theory \cite{goguen-burstall-1992}. The
pipeline includes the following phases:

>>>>

Nunchaku is a new higher-order model finder designed to be integrated with
several proof assistants. The first version was released in January 2016
with support for (co)al\-gebraic datatypes, (co)recursive functions, and
(co)inductive predicates. Support for higher-order functions is in the works.
We have developed an Isabelle frontend and are
planning further frontends for Coq, %HOL4, HOL Light,
the TLA\textsuperscript{+} Proof System, and other proof assistants.

Nunchaku is a spiritual successor to Nitpick \cite{blanchette-nipkow-2010} for
Isabelle/HOL, but is developed as a standalone
OCaml program, with its own input language. Whereas Nitpick generates a
succession of problems where cardinalities of finite types grow at each step,
Nunchaku translates its input to one first-order logic program that targets
the finite model finding fragment of CVC4, including
(co)algebraic datatypes \cite{reynolds-blanchette-2015-codata}.
Using CVC4 also allows Nunchaku to provide efficient arithmetic
reasoning and to detect unsatisfiability in addition to satisfiability.
We plan to integrate other tools as backends, to exploit the
strengths of competing approaches.
% and makes
% equality reasoning straightforward.

The input syntax was inspired by that other systems based on higher-order
logic (e.g., Isabelle/HOL) and by functional programming languages (e.g.,
OCaml). The following simple problem gives a taste of the syntax:
%
\begin{quote}
\verb$data nat := 0 | Suc nat.$ \\[2\jot]
\verb$pred even : nat -> prop :=$ \\
\verb$  even 0;$ \\
\verb$  forall n. odd n => even (Suc n)$ \\
\verb$and odd : nat -> prop :=$ \\
\verb$  forall n. even n => odd (Suc n).$ \\[2\jot]
\verb$val m : nat.$ \\
\verb$goal even m && ~ (m = 0).$
\end{quote}
%
The problem defines a datatype (\texttt{nat}) and two mutually recursive
inductive predicates (\texttt{even} and \texttt{odd}), declares a constant
\texttt{m}, and specifies a goal to satisfy (``\texttt{m} is even and nonzero'').
Nunchaku quickly finds the following partial model:
%
\begin{quote}
\verb$val m := Suc (Suc 0).$ \\
\verb$val odd := fun x. if x = Suc 0 then true else ?__.$ \\
\verb$val even := fun x. if x = Suc (Suc 0) || x = 0 then true else ?__.$\kern-200mm
\end{quote}
%
The partial model gives sufficient information to the user to evaluate the
goal: ``2 is even if 1 is odd, 1 is odd if 0 is even, and 0 is even.''
Our experience with Nitpick
is that users are mostly interested in the values assigned to uninterpreted
constants (e.g.,~\texttt{m}). Occasionally, the models of underspecified
recursive functions are instructive. A typical example is
the \texttt{head} function that returns the first element of a nonempty list:
%
\begin{quote}
\verb$data list A := Nil | Cons A (list A).$ \\[2\jot]
\verb$rec head : pi A. list A -> A :=$ \\
\verb$  forall y ys. head (Cons y ys) = y.$ \\[2\jot]
\verb$goal ~ (head Nil = 0).$
\end{quote}
%
Nunchaku transforms the definition of \verb$head$ into
%
\begin{quote}
\verb$head xs = match xs with Nil -> head xs | Cons y ys -> y end$
\end{quote}
%
where the unspecified \verb$Nil$ case is expressed via nonterminating
recursion (\verb$head$ \verb$xs$ \verb$=$ \verb$head$ \verb$xs$).
The tool exhibits a model in which \verb$head$ \verb$Nil$ is interpreted as
a nonzero value.

@@

Conceptually, the sequence of transformation is a \relax{bidirectional pipeline}
built by composing pairs $(\textit{Encode},\vthinspace \textit{Decode})$ of
transformations. For each such pair, \textit{Encode} translates a
$\Sigma$-problem in a logic~$\mathcal{L}$ to a $\Sigma'$-problem
in a logic~$\mathcal{L}'$, and \textit{Decode} translates a model
in~$\mathcal{L}'$ over $\Sigma'$ into a model in~$\mathcal{L}$ over $\Sigma$,
in the spirit of institution theory \cite{goguen-burstall-1992}. The
pipeline includes the following phases:
